$\pi$ with leftovers: \\ a mechanisation in Agda

### 2020-05-01 paper/paper.tex
Use autoref

\subsection{Contributions}
Our contributions are the following:
\begin{itemize}
\item \textbf{Formalisation of the syntax and the semantics of the \picalc{}}:
  Section \S \ref{syntax} uses type level de Bruijn indices \cite{} to introduce a syntax that is well-scoped by construction: every free variable is accounted for in the type of the process that uses it.
  Section \S \ref{semantics} provides an operational semantics for the untyped \picalc{}, prior to any typing.
  This operational semantics is given as a reduction relation on processes.
  The reduction relation tracks at the type level what channel the communication occurs on, so that this information can then be used to state subject reduction --- ak-a type preservation.
  It is also defined modulo structural congruence --- a relation that acts as a quotient type that gets rid of the unecessary syntactic minutiae introduced by the syntax of the \picalc{}.

  \item \textbf{Leftover typing for the \picalc{}, with linear, gradual and shared types}:
  Section \S \ref{type-system} provides the type system for a resource-aware \picalc{}.
  The type system is \emph{resource-aware} in that it allows the user to provide a resource-aware algebra that is then applied to the type-system.
  Linear types, gradual types and shared types are all instances of such a resource-aware algebra too.
  Any \emph{partial commutative monoid} that is cancellative and has a minimal element is a valid such algebra.
  Multiple of these algebras can be applied to a single type system --- contexts keep information about what algebra should be applied to which element.

@@

  \item \textbf{Machine verified proofs of subject reduction and auxiliary theorems}:
  Section \S \ref{type-safety} builds on several theorems to proof subject reduction for our type system.
  Here too, leftover typing renders the proofs for some of these lemmas straightforward, while for others it makes it possible to state them in the first place.
  We prove \textbf{framing} (well-typedness is independent of the unused resources of a process), \textbf{weakening} (new variables that are added to both the input and output usage context preserve well-typedness), and \textbf{strengthening} (removing the variables that are unused by a process preserves well-typedness).
  We use these to prove \textbf{subject congruence} (structural congruence rules can be freely applied while preserving well-typedness) and \textbf{substitution} (a variable can be substituted for another as long as this other variable is capable of doing everything the original one did), and use these to prove \textbf{subject reduction} (a well-typed process that reduces will still be well-typed).

@@

\begin{figure}[h]
  \begin{mathpar}
    \datatype
    { }
    {\type{\N} : \Set}
    \; \textsc{Nat}

@@

Data type definitions use double lines and index-free synonyms as rule names (for ease of reference).
We otherwise use the constructor name as the name of a typing rule.
Universe levels and universe polymorphism are omitted for brevity --- all our types are of type $\Set$.
Implicit arguments are mentioned by type definitions but omitted in constructors.
$\type{TYPES}$ are in blue and uppercased (with indices as subscripts), $\constr{constructors}$ are in olive, $\field{fields}$ in red, and variables and functions in black.
Some constructor names are overloaded --- they and disambiguated by context.

@@

\begin{figure}[h]
\begin{mathpar}
  \datatype
  {n : \N}
  {\Var_n : \Set}
  \; \textsc{Var}

@@

Every time we go under a binder, the number of binders a variable might refer to increments by one.
To propagate this information, we index processes according to their \emph{depth}: for all naturals $n$, a process of type $\Process_n$ contains variables that can refer to $n$ distinct elements.
Figure \ref{fig:syntax} shows the syntax of \picalc{} in Agda. We increase the \emph{depth} counter by 1 in channel restriction and input processes.

\begin{figure}[h]
\begin{mathpar}
  \datatype
  {n : \N}
  {\Process_n : \Set}
  \; \textsc{Process}
\end{mathpar}

@@

Structural congruence is a congruent equivalence relation on processes.
Figure \ref{fig:struct-cong-base} lists the base cases of structural congruence.
The type $\Unused_i \; Q$ is an inductive proof that witnesses that the variable index $i$ does not appear nor in the inputs nor in the outputs of $Q$.
The function $lower_i \; Q \; uQ$ traverses $Q$ decrementing every index bigger than $i$.
Finally, $swap_i \; P$ traverses $P$ (of type $\Process_{\suc \suc n}$) and swaps variable references $i$ and $\suc i$.
In all three, $i$ is incremented every time we go under a binder.

\begin{figure}[h]
\begin{mathpar}
  \datatype
  { }
  {P \eqeq Q : \Set}
  \; \textsc{StructCong}

@@

Structural congruence is a congruent equivalence relation.
As such, rewrites can happen anywhere inside a process, and are closed under reflexivity, symmetry and transitivity as shown in Figure \ref{fig:struct-cong}.
In \S \ref{subject-congruence} we prove that if two processes $P$ and $Q$ are structurally congruent and $P$ is well-typed, then so is $Q$.
In the transitivity case, we must show that if $P$ is structurally congruent to $Q$ and $Q$ is to $R$, and $P$ is well-typed, then so is $R$.
To do so, we need to proceed by induction and first get a proof of the well-typedness of $Q$, then use that to reach $R$.
To show the typechecker that the doubly recursive call terminates we index the equivalence relation $=$ by a type $\Rec$ that models the structure of the recursion.

\begin{figure}[h]
\begin{mathpar}
  \datatype
  { }
  {\Rec : \Set}
  \; \textsc{Rec}

@@

  \datatype
  {P \, Q : \Process_n \\ r : \Rec}
  {P \eq{r} Q : \Set}
  \; \textsc{Equals}

@@

Figure \ref{fig:reduction} models the operational semantics of the \picalc{}.
Processes put in parallel reduce when they communicate over a common channel.
The receiving process substitutes references to its most immediate variable with references to the variable that is being sent by the process doing the output.
The process is then \emph{lowered} --- all variable references are decreased by one.
Reduction is closed under structural congruence and goes under parallel composition and scope restriction
--- notably, not under input nor output, or otherwise the sequencing of actions would not be preserved.
In \S \ref{subject-reduction} we prove that if $P$ reduces to $Q$ and $P$ is well-typed, so is $Q$.
However, this reduction operation is effectful: it consumes the channel over which communication occurs.
If this channel is external to $P$ (it resides in its context) then the context in which $Q$ is typed must change.
To keep track of this information, we lift the index of the channel to the type level.
We use a $dec$ function to decrement the index of the channel every time we come out of a binder.
This function saturates at $\constr{internal}$ --- meaning communication happens internally to the process.

\begin{figure}[h]
\begin{mathpar}
  \datatype
  {n : \N}
  {\Channel_n : \Set}
  \; \textsc{Channel}

@@

  \datatype
  {i : \Channel_n \\ P \; Q : \Process_n}
  {P \reduce{i} Q : \Set}
  \; \textsc{Reduces}

@@

Typing systems for the \picalc{} ensure that processes make a correct use of their channels.
Typing judgments are of the form $\Gamma \vdash P$ (for some typing context $\Gamma$ and some process $P$) --- types only appear in the contexts of typing judgments.
A resource-aware typing system for the \picalc{} must take resource usages into account, and needs an algebra on these --- a way of \emph{splitting} resources on parallel composition, of \emph{ticking off} resources on input and output.
We characterise such an algebra in \S \ref{multiplicities}.
We define independent typing contexts and usage contexts in \S \ref{contexts}.
We provide a typing judgment for a resource-aware \picalc{} in \S \ref{leftover-typing}.

@@

\todo{we haven't introduced the forall and implicit syntax}
\begin{nidefinition}
  The algebra on multiplicities must be \emph{partial}, \emph{functional}, \emph{cancelative}, \emph{associtive}, and \emph{commutative}, and needs a \emph{minimal neutral element}.
  It also requires a special element that is to be used to count inputs and outputs.
  We define the laws of such a \emph{partial commutative monoid} as a type $\Quantifier_C$ on a carrier set $C$ in Figure \ref{fig:multiplicities} --- we use a terniary relation to model the monoidal operation.

  \begin{figure}[h]
  \begin{equation}
  \begin{aligned}
    &\zero                  &:{} &                      &        & C \\
    &\one                   &:{} &                      &        & C \\
    &\op{\_}{\_}{\_}        &:{} &                      &        & C \to C \to C \to \Set \\
    &\field{\cdot-compute}  &:{} &\forall y z           & \to \; & \type{DEC} \; (\type{\exists} x . \; (\op{x}{y}{z})) \\
    &\field{\cdot-unique}   &:{} &\forall \{x x' y z\}  & \to \; & \op{x'}{y}{z} \to \op{x}{y}{z} \to x' \equiv x \\
    &\field{\cdot-unique^l} &:{} &\forall \{x y y' z\}  & \to \; & \op{x}{y'}{z} \to \op{x}{y}{z} \to y' \equiv y \\
    &\field{0\cdot-min^l}   &:{} &\forall \{y z\}       & \to \; & \op{\zero}{y}{z} \to y \equiv \zero \\
    &\field{\cdot-id^l}     &:{} &\forall \{x\}         & \to \; & \op{x}{\zero}{x} \\
    &\field{\cdot-comm}     &:{} &\forall \{x y z\}     & \to \; & \op{x}{y}{z} \to \op{x}{z}{y} \\
    &\field{\cdot-assoc}    &:{} &\forall \{x y z u v\} & \to \; & \op{x}{y}{z} \to \op{y}{u}{v} \to \type{\exists} w . \; (\op{x}{u}{w} \times \op{w}{v}{z})
  \end{aligned}
  \end{equation}
  \caption{Quantifier algebra $\Quantifier_C$ algebra on a partial commutative monoid.}
  \label{fig:multiplicities}
  \end{figure}
\end{nidefinition}

@@

Shared, gradual and linear multiplicity types can all be defined as an instance of the properties in Figure \ref{fig:multiplicities}.
For examples of how these multiplicity types are used in typing derivations, refer to Example \ref{example-derivations}.

@@

\begin{nidefinition}
  We define types (Figure \ref{fig:prectx}) as either unit types, base types or channel types.
  The unit type $\unit$ serves as a proof of inhabitance for types (we can trivially obtain it), and is used to such end in subject congruence --- for the $\new \PO_{\suc n} \eqeq \PO_n$ rule.
  The base type $\base{n}$ uses a natural number $n$ to represent types.
  This natural number can then be interpreted as a type in the host language.
  Using natural numbers instead of type universes is convenient: less universe polymorphism poluting specifications.

@@

  \begin{figure}[h]
  \begin{mathpar}
    \datatype
    { }
    {\Type : \Set}
    \; \textsc{Type}

@@

  \begin{mathpar}
    \datatype
    {n : \N}
    {\PreCtx_n : \Set}
    \; \textsc{PreCtx}

@@

\begin{nidefinition}
  We use indexed sets of usage algebras to allow several usage algebras to coexist in our type system.
  An indexed set of usage algebras is a set of indices such that for every index there exists a carrier $C$ with a $\Quantifier_C$ algebra defined on it, as captured by Figure \ref{fig:indexed-multiplicities}.
  We require the index type to be inhabited -- proving subject congruence on $\new \PO_{\suc n} \eqeq \PO_n$ requires an inhabited carrier set.

  \begin{figure}[h]
  \begin{equation}
  \begin{split}
    &\field{IDX}          &:{} &\Set \\
    &\field{\exists IDX}  &:{} &\field{IDX} \\
    &\field{CARRIER}      &:{} &\field{IDX} \to \Set \\
    &\field{QUANTIFIERS}  &:{} &(i : \field{IDX}) \to \Quantifier_{\field{CARRIER}_i}
  \end{split}
  \end{equation}
  \caption{Indexed set of partial commutative monoids.}
  \label{fig:indexed-multiplicities}
  \end{figure}
\end{nidefinition}

\begin{nidefinition}
  A context of usages $\Ctx_{idxs}$ is indexed by a length-indexed context of indices $\Idxs_n$, as shown in Figure \ref{fig:ctx}.
  \begin{figure}[h]
  \begin{mathpar}
    \datatype
    {n : \N}
    {\Idxs_n : \Set}
    \; \textsc{Idxs}

@@

    \datatype
    {idxs : \Idxs_n}
    {\Ctx_{idxs} : \Set}
    \; \textsc{Ctx}

@@

We make use \emph{leftover typing} \cite{}, an approach that, in addition to the usual typing context and input usage context, adds an extra output usage context to the typing rules.
This output context contains the \emph{leftovers} of the process being typed --- its unused multiplicities.
Crucially, the leftovers can then be used as an input to some other typing derivation.
Adopting this techinque has three main advantages:
\begin{itemize}
  \item
    The input and output actions no longer transform the usage context through functions.
    Instead, they use a variable reference type with both an input and an output usage context that encapsulates these transformations.
  \item
    Parallel composition no longer needs an explicit context splitting proof --- a proof $\opctx{\Gamma_m}{\Gamma_l}{\Gamma_r}$ where $\Gamma_l \vdash P$, $\Gamma_r \vdash Q$ and $\Gamma_m \vdash \comp{P}{Q}$.
    Such a proof could be obtained through automated search --- but it is better yet to avoid needing it.
    With leftover typing, we use the leftovers of $P$ to type $Q$.
  \item
    Theorems like framing (\S \ref{framing}), weakening (\S \ref{weakening}) and strengthening (\S \ref{strengthening}) become suddenly stateable.
\end{itemize}

We introduce two judgments: one for variable references (Definition \ref{def:varref}), one for processes (Definition \ref{def:types}).
Both of them combine a judgment on types and a judgment on usages.
These could be kept separate, but doing so would risk incoherence: if typing derivations had any choice to make, the judgment on types could choose to do one thing while the judgment on usages chooses to do another.
The constructions instroduced in this paper make the typing derivations completely syntax directed and render this problem inexistent, but we nevertheless choose to keep both judgments as one, for simplicity.

@@

  The \textsc{VarRef} relation presented in Figure \ref{fig:variable-references} is indexed by a typing context and an input and output usage context.
  In addition to these, it also keeps track of its position in the contexts, its type, and its usage annotation.

@@

    \datatype{
      \gamma : \PreCtx_n \\
      \stacked{
        idxs : \Idxs_n \\\\
        \Gamma : \Ctx_{idxs}} \\
      i : \Var_n \\
      t : \Type \\
      \stacked{
        idx : \Idx \\\\
        y : \Carrier_{idx}^2} \\
      \Delta : \Ctx_{idxs}}
    {\contains{\gamma}{\Gamma}{i}{t}{y}{\Delta} : \Set}
    \; \textsc{VarRef}

@@

  The typing relation \textsc{Types} presented in Figure \ref{fig:types} is indexed by a typing context, an input and an output usage context, and the process $P$ being typed.

  \begin{figure}[h]
  \begin{mathpar}
    \datatype{
      \gamma : \PreCtx_n \\
      \stacked{
        idxs : \Idxs_n \\\\
        \Gamma : \Ctx_{idxs}} \\
      P : \Process_n \\
      \Delta : \Ctx_{idxs}}
    {\types{\gamma}{\Gamma}{P}{\Delta} : \Set}
    \; \textsc{Types}

@@

\begin{nitheorem}
  \label{thm:framing}
  Let $P$ be a well-typed process from context $\Gamma_l$ to $\Xi_l$.
  Then, per Lemma \ref{lm:types-op}, there exists some $\Delta$ such that $\opctx{\Gamma_l}{\Delta}{\Xi_l}$.
  Take arbitrary contexts $\Gamma_r$ and $\Xi_r$ such that $\opctx{\Gamma_r}{\Delta}{\Xi_r}$.
  Then $\types{\gamma}{\Gamma_r}{P}{\Xi_r}$.
\end{nitheorem}
\begin{proof}
  By induction on \textsc{Process} and \textsc{Var}.
\end{proof}

@@

\begin{proof}
  By induction on \textsc{Equals}.
  The case $\constr{struct}$ and its symmetric one under $\constr{sym}$ are the nontrivial ones.
  For those, we proceed by induction on \textsc{StructCong}:
  \begin{itemize}
    \item
      The case $\constr{comp-assoc}$ is trivial because the encoding with leftover typing of the typing rules naturally preserves associativity.
    \item
      The $\constr{comp-sym}$ case for a process $\comp{P}{Q}$ is however more complicated: $P$ must use framing (Theorem \ref{thm:framing}) to shift its output context to $Q$'s; $Q$ must use framing to shift its input context to $P$'s.
      The usage context on which $Q$ and $P$ coincide in $\comp{Q}{P}$ will no longer be the same.
    \item
      The case $\constr{comp-end}$ is trivial because in the typing rule for process $\PO$ the input and output contexts are one and the same.
    \item
      In the $\constr{scope-end}$ case, the typing rules $\constr{chan}$ and $\constr{end}$ work together to constraint the multiplicity of the newly create channel to $\lz$, making the proof trivial.
      In the opposite direction, we will need to instantiate the newly created channel to a concrete type and multiplicities.
      For this, we will use type $\unit$ and multiplicities $\lz$.
    \item
      In the $\constr{scope-ext}$ case for a process $\new \comp{P}{Q}$, we will make use of Lemma $\ref{lm:types-unused}$ to show that as the typing derivation on $P$ will preserved the multiplicity of the recently introduced variable unchanged.
      We then use strengthening (Theorem \ref{thm:strengthening}) to delete the recently introduced variable from $P$, and then compose with the typing derivation of $Q$ under $\constr{chan}$.
      In the reverse direction, we will weaken $P$ so that it can accept an additional variable.
      We will then use Lemma \ref{lm:lower-lift} to show that lifting a lowered $P$ leaves $P$ unchanged.
    \item
      Finally, for the $\constr{scope-comm}$ rule, we will use swapping as shown in Theorem \ref{thm:swapping}.
      In the reverse direction, we will use swapping and then use Lemma \ref{lm:swap-swap} to show that swapping indices twice leaves $P$ unchanged.
  \end{itemize}
\end{proof}
\todo{make sure constructor names in code coincide}

@@

Subject reduction (aka type preservation) relates the operational semantics and the typing system of a language.
In its usual form, it states that if $P$ reduces to $Q$ and $P$ is well-typed under context $\Gamma$, then $Q$ is well-typed under $\Gamma$ too.
In the linear \picalc{} however, reduction \emph{is} meant to transform the context: each time communication occurs over a channel, that channel is exhausted.
In our case, and because we generalize linearity to a partial monoid, the channel over which the communication occurs will loose a multiplicity $\one$ on both its input and output capabilities.
In \S \ref{semantics}, we made sure to lift the index of the communicating channel to the type level, now we use it to state subject reduction.

>>>>

\subsection{Contributions}
Our contributions are the following:
\begin{itemize}
\item \textbf{Formalisation of the syntax and the semantics of the \picalc{}}:
  \autoref{syntax} uses type level de Bruijn indices \cite{} to introduce a syntax that is well-scoped by construction: every free variable is accounted for in the type of the process that uses it.
  \autoref{semantics} provides an operational semantics for the untyped \picalc{}, prior to any typing.
  This operational semantics is given as a reduction relation on processes.
  The reduction relation tracks at the type level what channel the communication occurs on, so that this information can then be used to state subject reduction --- ak-a type preservation.
  It is also defined modulo structural congruence --- a relation that acts as a quotient type that gets rid of the unecessary syntactic minutiae introduced by the syntax of the \picalc{}.

  \item \textbf{Leftover typing for the \picalc{}, with linear, gradual and shared types}:
  \autoref{type-system} provides the type system for a resource-aware \picalc{}.
  The type system is \emph{resource-aware} in that it allows the user to provide a resource-aware algebra that is then applied to the type-system.
  Linear types, gradual types and shared types are all instances of such a resource-aware algebra too.
  Any \emph{partial commutative monoid} that is cancellative and has a minimal element is a valid such algebra.
  Multiple of these algebras can be applied to a single type system --- contexts keep information about what algebra should be applied to which element.

@@

  \item \textbf{Machine verified proofs of subject reduction and auxiliary theorems}:
  \autoref{type-safety} builds on several theorems to proof subject reduction for our type system.
  Here too, leftover typing renders the proofs for some of these lemmas straightforward, while for others it makes it possible to state them in the first place.
  We prove \textbf{framing} (well-typedness is independent of the unused resources of a process), \textbf{weakening} (new variables that are added to both the input and output usage context preserve well-typedness), and \textbf{strengthening} (removing the variables that are unused by a process preserves well-typedness).
  We use these to prove \textbf{subject congruence} (structural congruence rules can be freely applied while preserving well-typedness) and \textbf{substitution} (a variable can be substituted for another as long as this other variable is capable of doing everything the original one did), and use these to prove \textbf{subject reduction} (a well-typed process that reduces will still be well-typed).

@@

\begin{figure}[h]
  \begin{mathpar}
    \datatype
    { }
    {\type{\N} : et}
    \; \textsc{Nat}

@@

Data type definitions use double lines and index-free synonyms as rule names (for ease of reference).
We otherwise use the constructor name as the name of a typing rule.
Universe levels and universe polymorphism are omitted for brevity --- all our types are of type $et$.
Implicit arguments are mentioned by type definitions but omitted in constructors.
$\type{TYPES}$ are in blue and uppercased (with indices as subscripts), $\constr{constructors}$ are in olive, $\field{fields}$ in red, and variables and functions in black.
Some constructor names are overloaded --- they and disambiguated by context.

@@

\begin{figure}[h]
\begin{mathpar}
  \datatype
  {n : \N}
  {\Var_n : et}
  \; \textsc{Var}

@@

Every time we go under a binder, the number of binders a variable might refer to increments by one.
To propagate this information, we index processes according to their \emph{depth}: for all naturals $n$, a process of type $\Process_n$ contains variables that can refer to $n$ distinct elements.
\autoref{fig:syntax} shows the syntax of \picalc{} in Agda. We increase the \emph{depth} counter by 1 in channel restriction and input processes.

\begin{figure}[h]
\begin{mathpar}
  \datatype
  {n : \N}
  {\Process_n : et}
  \; \textsc{Process}
\end{mathpar}

@@

Structural congruence is a congruent equivalence relation on processes.
\autoref{fig:struct-cong-base} lists the base cases of structural congruence.
The type $\Unused_i \; Q$ is an inductive proof that witnesses that the variable index $i$ does not appear nor in the inputs nor in the outputs of $Q$.
The function $lower_i \; Q \; uQ$ traverses $Q$ decrementing every index bigger than $i$.
Finally, $swap_i \; P$ traverses $P$ (of type $\Process_{\suc \suc n}$) and swaps variable references $i$ and $\suc i$.
In all three, $i$ is incremented every time we go under a binder.

\begin{figure}[h]
\begin{mathpar}
  \datatype
  { }
  {P \eqeq Q : et}
  \; \textsc{StructCong}

@@

Structural congruence is a congruent equivalence relation.
As such, rewrites can happen anywhere inside a process, and are closed under reflexivity, symmetry and transitivity as shown in \autoref{fig:struct-cong}.
In  \autoref{subject-congruence} we prove that if two processes $P$ and $Q$ are structurally congruent and $P$ is well-typed, then so is $Q$.
In the transitivity case, we must show that if $P$ is structurally congruent to $Q$ and $Q$ is to $R$, and $P$ is well-typed, then so is $R$.
To do so, we need to proceed by induction and first get a proof of the well-typedness of $Q$, then use that to reach $R$.
To show the typechecker that the doubly recursive call terminates we index the equivalence relation $=$ by a type $\Rec$ that models the structure of the recursion.

\begin{figure}[h]
\begin{mathpar}
  \datatype
  { }
  {\Rec : et}
  \; \textsc{Rec}

@@

  \datatype
  {P \, Q : \Process_n \\ r : \Rec}
  {P \eq{r} Q : et}
  \; \textsc{Equals}

@@

\autoref{fig:reduction} models the operational semantics of the \picalc{}.
Processes put in parallel reduce when they communicate over a common channel.
The receiving process substitutes references to its most immediate variable with references to the variable that is being sent by the process doing the output.
The process is then \emph{lowered} --- all variable references are decreased by one.
Reduction is closed under structural congruence and goes under parallel composition and scope restriction
--- notably, not under input nor output, or otherwise the sequencing of actions would not be preserved.
In  \autoref{subject-reduction} we prove that if $P$ reduces to $Q$ and $P$ is well-typed, so is $Q$.
However, this reduction operation is effectful: it consumes the channel over which communication occurs.
If this channel is external to $P$ (it resides in its context) then the context in which $Q$ is typed must change.
To keep track of this information, we lift the index of the channel to the type level.
We use a $dec$ function to decrement the index of the channel every time we come out of a binder.
This function saturates at $\constr{internal}$ --- meaning communication happens internally to the process.

\begin{figure}[h]
\begin{mathpar}
  \datatype
  {n : \N}
  {\Channel_n : et}
  \; \textsc{Channel}

@@

  \datatype
  {i : \Channel_n \\ P \; Q : \Process_n}
  {P \reduce{i} Q : et}
  \; \textsc{Reduces}

@@

Typing systems for the \picalc{} ensure that processes make a correct use of their channels.
Typing judgments are of the form $\Gamma \vdash P$ (for some typing context $\Gamma$ and some process $P$) --- types only appear in the contexts of typing judgments.
A resource-aware typing system for the \picalc{} must take resource usages into account, and needs an algebra on these --- a way of \emph{splitting} resources on parallel composition, of \emph{ticking off} resources on input and output.
We characterise such an algebra in  \autoref{multiplicities}.
We define independent typing contexts and usage contexts in  \autoref{contexts}.
We provide a typing judgment for a resource-aware \picalc{} in  \autoref{leftover-typing}.

@@

\todo{we haven't introduced the forall and implicit syntax}
\begin{nidefinition}
  The algebra on multiplicities must be \emph{partial}, \emph{functional}, \emph{cancelative}, \emph{associtive}, and \emph{commutative}, and needs a \emph{minimal neutral element}.
  It also requires a special element that is to be used to count inputs and outputs.
  We define the laws of such a \emph{partial commutative monoid} as a type $\Quantifier_C$ on a carrier set $C$ in \autoref{fig:multiplicities} --- we use a terniary relation to model the monoidal operation.

  \begin{figure}[h]
  \begin{equation}
  \begin{aligned}
    &\zero                  &:{} &                      &        & C \\
    &\one                   &:{} &                      &        & C \\
    &\op{\_}{\_}{\_}        &:{} &                      &        & C \to C \to C \to et \\
    &\field{\cdot-compute}  &:{} &\forall y z           & \to \; & \type{DEC} \; (\type{\exists} x . \; (\op{x}{y}{z})) \\
    &\field{\cdot-unique}   &:{} &\forall \{x x' y z\}  & \to \; & \op{x'}{y}{z} \to \op{x}{y}{z} \to x' \equiv x \\
    &\field{\cdot-unique^l} &:{} &\forall \{x y y' z\}  & \to \; & \op{x}{y'}{z} \to \op{x}{y}{z} \to y' \equiv y \\
    &\field{0\cdot-min^l}   &:{} &\forall \{y z\}       & \to \; & \op{\zero}{y}{z} \to y \equiv \zero \\
    &\field{\cdot-id^l}     &:{} &\forall \{x\}         & \to \; & \op{x}{\zero}{x} \\
    &\field{\cdot-comm}     &:{} &\forall \{x y z\}     & \to \; & \op{x}{y}{z} \to \op{x}{z}{y} \\
    &\field{\cdot-assoc}    &:{} &\forall \{x y z u v\} & \to \; & \op{x}{y}{z} \to \op{y}{u}{v} \to \type{\exists} w . \; (\op{x}{u}{w} \times \op{w}{v}{z})
  \end{aligned}
  \end{equation}
  \caption{Quantifier algebra $\Quantifier_C$ algebra on a partial commutative monoid.}
  \label{fig:multiplicities}
  \end{figure}
\end{nidefinition}

@@

Shared, gradual and linear multiplicity types can all be defined as an instance of the properties in \autoref{fig:multiplicities}.
For examples of how these multiplicity types are used in typing derivations, refer to Example \autoref{example-derivations}.

@@

\begin{nidefinition}
  We define types (\autoref{fig:prectx}) as either unit types, base types or channel types.
  The unit type $\unit$ serves as a proof of inhabitance for types (we can trivially obtain it), and is used to such end in subject congruence --- for the $\new \PO_{\suc n} \eqeq \PO_n$ rule.
  The base type $\base{n}$ uses a natural number $n$ to represent types.
  This natural number can then be interpreted as a type in the host language.
  Using natural numbers instead of type universes is convenient: less universe polymorphism poluting specifications.

@@

  \begin{figure}[h]
  \begin{mathpar}
    \datatype
    { }
    {\Type : et}
    \; \textsc{Type}

@@

  \begin{mathpar}
    \datatype
    {n : \N}
    {\PreCtx_n : et}
    \; \textsc{PreCtx}

@@

\begin{nidefinition}
  We use indexed sets of usage algebras to allow several usage algebras to coexist in our type system.
  An indexed set of usage algebras is a set of indices such that for every index there exists a carrier $C$ with a $\Quantifier_C$ algebra defined on it, as captured by \autoref{fig:indexed-multiplicities}.
  We require the index type to be inhabited -- proving subject congruence on $\new \PO_{\suc n} \eqeq \PO_n$ requires an inhabited carrier set.

  \begin{figure}[h]
  \begin{equation}
  \begin{split}
    &\field{IDX}          &:{} &et \\
    &\field{\exists IDX}  &:{} &\field{IDX} \\
    &\field{CARRIER}      &:{} &\field{IDX} \to et \\
    &\field{QUANTIFIERS}  &:{} &(i : \field{IDX}) \to \Quantifier_{\field{CARRIER}_i}
  \end{split}
  \end{equation}
  \caption{Indexed set of partial commutative monoids.}
  \label{fig:indexed-multiplicities}
  \end{figure}
\end{nidefinition}

\begin{nidefinition}
  A context of usages $\Ctx_{idxs}$ is indexed by a length-indexed context of indices $\Idxs_n$, as shown in \autoref{fig:ctx}.
  \begin{figure}[h]
  \begin{mathpar}
    \datatype
    {n : \N}
    {\Idxs_n : et}
    \; \textsc{Idxs}

@@

    \datatype
    {idxs : \Idxs_n}
    {\Ctx_{idxs} : et}
    \; \textsc{Ctx}

@@

We make use \emph{leftover typing} \cite{}, an approach that, in addition to the usual typing context and input usage context, adds an extra output usage context to the typing rules.
This output context contains the \emph{leftovers} of the process being typed --- its unused multiplicities.
Crucially, the leftovers can then be used as an input to some other typing derivation.
Adopting this techinque has three main advantages:
\begin{itemize}
  \item
    The input and output actions no longer transform the usage context through functions.
    Instead, they use a variable reference type with both an input and an output usage context that encapsulates these transformations.
  \item
    Parallel composition no longer needs an explicit context splitting proof --- a proof $\opctx{\Gamma_m}{\Gamma_l}{\Gamma_r}$ where $\Gamma_l \vdash P$, $\Gamma_r \vdash Q$ and $\Gamma_m \vdash \comp{P}{Q}$.
    Such a proof could be obtained through automated search --- but it is better yet to avoid needing it.
    With leftover typing, we use the leftovers of $P$ to type $Q$.
  \item
    Theorems like framing ( \autoref{framing}), weakening ( \autoref{weakening}) and strengthening ( \autoref{strengthening}) become suddenly stateable.
\end{itemize}

We introduce two judgments: one for variable references (\autoref{def:varref}), one for processes (\autoref{def:types}).
Both of them combine a judgment on types and a judgment on usages.
These could be kept separate, but doing so would risk incoherence: if typing derivations had any choice to make, the judgment on types could choose to do one thing while the judgment on usages chooses to do another.
The constructions instroduced in this paper make the typing derivations completely syntax directed and render this problem inexistent, but we nevertheless choose to keep both judgments as one, for simplicity.

@@

  The \textsc{VarRef} relation presented in \autoref{fig:variable-references} is indexed by a typing context and an input and output usage context.
  In addition to these, it also keeps track of its position in the contexts, its type, and its usage annotation.

@@

    \datatype{
      \gamma : \PreCtx_n \\
      \stacked{
        idxs : \Idxs_n \\\\
        \Gamma : \Ctx_{idxs}} \\
      i : \Var_n \\
      t : \Type \\
      \stacked{
        idx : \Idx \\\\
        y : \Carrier_{idx}^2} \\
      \Delta : \Ctx_{idxs}}
    {\contains{\gamma}{\Gamma}{i}{t}{y}{\Delta} : et}
    \; \textsc{VarRef}

@@

  The typing relation \textsc{Types} presented in \autoref{fig:types} is indexed by a typing context, an input and an output usage context, and the process $P$ being typed.

  \begin{figure}[h]
  \begin{mathpar}
    \datatype{
      \gamma : \PreCtx_n \\
      \stacked{
        idxs : \Idxs_n \\\\
        \Gamma : \Ctx_{idxs}} \\
      P : \Process_n \\
      \Delta : \Ctx_{idxs}}
    {\types{\gamma}{\Gamma}{P}{\Delta} : et}
    \; \textsc{Types}

@@

\begin{nitheorem}
  \label{thm:framing}
  Let $P$ be a well-typed process from context $\Gamma_l$ to $\Xi_l$.
  Then, per \autoref{lm:types-op}, there exists some $\Delta$ such that $\opctx{\Gamma_l}{\Delta}{\Xi_l}$.
  Take arbitrary contexts $\Gamma_r$ and $\Xi_r$ such that $\opctx{\Gamma_r}{\Delta}{\Xi_r}$.
  Then $\types{\gamma}{\Gamma_r}{P}{\Xi_r}$.
\end{nitheorem}
\begin{proof}
  By induction on \textsc{Process} and \textsc{Var}.
\end{proof}

@@

\begin{proof}
  By induction on \textsc{Equals}.
  The case $\constr{struct}$ and its symmetric one under $\constr{sym}$ are the nontrivial ones.
  For those, we proceed by induction on \textsc{StructCong}:
  \begin{itemize}
    \item
      The case $\constr{comp-assoc}$ is trivial because the encoding with leftover typing of the typing rules naturally preserves associativity.
    \item
      The $\constr{comp-sym}$ case for a process $\comp{P}{Q}$ is however more complicated: $P$ must use framing (\autoref{thm:framing}) to shift its output context to $Q$'s; $Q$ must use framing to shift its input context to $P$'s.
      The usage context on which $Q$ and $P$ coincide in $\comp{Q}{P}$ will no longer be the same.
    \item
      The case $\constr{comp-end}$ is trivial because in the typing rule for process $\PO$ the input and output contexts are one and the same.
    \item
      In the $\constr{scope-end}$ case, the typing rules $\constr{chan}$ and $\constr{end}$ work together to constraint the multiplicity of the newly create channel to $\lz$, making the proof trivial.
      In the opposite direction, we will need to instantiate the newly created channel to a concrete type and multiplicities.
      For this, we will use type $\unit$ and multiplicities $\lz$.
    \item
      In the $\constr{scope-ext}$ case for a process $\new \comp{P}{Q}$, we will make use of $\autoref{lm:types-unused}$ to show that as the typing derivation on $P$ will preserved the multiplicity of the recently introduced variable unchanged.
      We then use strengthening (\autoref{thm:strengthening}) to delete the recently introduced variable from $P$, and then compose with the typing derivation of $Q$ under $\constr{chan}$.
      In the reverse direction, we will weaken $P$ so that it can accept an additional variable.
      We will then use \autoref{lm:lower-lift} to show that lifting a lowered $P$ leaves $P$ unchanged.
    \item
      Finally, for the $\constr{scope-comm}$ rule, we will use swapping as shown in \autoref{thm:swapping}.
      In the reverse direction, we will use swapping and then use \autoref{lm:swap-swap} to show that swapping indices twice leaves $P$ unchanged.
  \end{itemize}
\end{proof}
\todo{make sure constructor names in code coincide}

@@

Subject reduction (aka type preservation) relates the operational semantics and the typing system of a language.
In its usual form, it states that if $P$ reduces to $Q$ and $P$ is well-typed under context $\Gamma$, then $Q$ is well-typed under $\Gamma$ too.
In the linear \picalc{} however, reduction \emph{is} meant to transform the context: each time communication occurs over a channel, that channel is exhausted.
In our case, and because we generalize linearity to a partial monoid, the channel over which the communication occurs will loose a multiplicity $\one$ on both its input and output capabilities.
In  \autoref{semantics}, we made sure to lift the index of the communicating channel to the type level, now we use it to state subject reduction.
